### 2017-01-04 Chapters/theory.tex
Layout fix, changing size of some figures

\begin{figure}[tb]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig4}
\caption{
\textbf{Optimal control of the self-replicator during a nutrient upshift.}
\textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines. The optimal trajectory is shown as a solid, red curve.
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the switching curve $\varphi(\hat{p})$.
The optimal solution was obtained by numerical optimization using \texttt{bocop}~\cite{bonnans_bocop_2012} (see \textit{Methods} for details), using the parameter values $E_M=1$ and $K=0.003$, and starting from the initial state $(0.024,0.18)$ at $t=0$ (optimal steady state for $E_M=0.2$).
\textit{B:}~Time evolution of the control variable $\alpha_{opt}(\cdot)$ (thick, red line) and the environment $E_M$ (dashed, black line).
}
\label{fig:optimalcontrol}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/Fig6}
\caption{
\textbf{Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}
\textit{A:}~Trajectory in the phase plane for the nutrient-only strategy (green curve).
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the $\hat{r}$-nullcline.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with $\alpha = f(E_M)$ as specified by Eq.~\ref{eq:meth_f} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The initial state corresponds to the steady state attained for an environment given by $0.2 \, E_M$.
While converging to the new steady state after the upshift, the precursor concentration makes a large overshoot.
\textit{B:}~As above, but for the precursor-only strategy.
The feedback control strategy is now defined by $\alpha = g(\hat{p})$ as specified by Eq.~\ref{eq:meth_g} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The solution trajectory (blue curve) exhibits a lower overshoot.
\textit{C:}~Evolution of the control variable $\alpha(\cdot)$ as a function of time, for each of the above two strategies.
Notice that in the nutrient-only strategy $\alpha(\cdot)$ immediately jumps to the optimal value for the post-upshift steady state (green curve), whereas in the precursor-only strategy it depends on the (time-varying) precursor concentration (blue curve).
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$  as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:comparison_nutrients_precursor}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/Fig7}
\caption{
\textbf{Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}
\textit{A:}~Trajectory in the phase plane for the on-off strategy (yellow curve).
The solid, black curve represents the $\hat{p}$-nullcline and the dashed, black curve the function $g$.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with the equation $\alpha = h(\hat{p},\hat{r})$ defined in Eq.~\ref{eq:stratswitch} and plotted in Fig.~\ref{fig:ppGppsurface}\textit{A}.
The initial state corresponds to the optimal steady state attained for an environment given by $0.2 \, E_M$.
\textit{B:}~Trajectory in the phase plane for the precursor-only strategy (same as in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{B}, added for comparison).
\textit{C:}~Evolution of the control variable $\alpha$ for each strategy as a function of time.
Both strategies stabilize the system at the optimal steady state, but only the on-off strategy (yellow curve) exhibits bang-bang behavior.
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$ for the on-off and precursor-only strategies as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
The final values of $\texttt{Vol} / \texttt{Vol}_{opt}$ attained by the two strategies are 0.9831 and 0.9413, respectively.
The on-off strategy is thus hardly distinguishable from the optimal control strategy in the plot.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:onoffresults}
\end{figure}

@@

As stated in the section \textit{Biomass maximization as an optimal control problem} in the main text, the objective of this study is to maximize the growth rate on an interval $[0,\tau]$ after a nutrient upshift. With Eq.~\ref{S1.eq:supp_growthrate_adim} in \ref{S1_Text}, we have

@@

An extremal trajectory is a quadruplet $(\hat{x}(\cdot),\lambda(\cdot),\lambda_0,\alpha(\cdot))$ satisfying Eqs~\ref{eq:model}-\ref{eq:PMP}. The extremal is said to be normal (resp. abnormal) if $\lambda_0<0$ (resp. $\lambda_0=0$). In the normal case, we normalize the adjoint vector so that $\lambda_0=-1$.\\

@@

A necessary condition of optimality for a singular arc is given by the Kelley condition~\cite{borisov_fullers_2000}.
We must differentiate $\phi$ with respect to $\hat{t}$ until $\alpha$ appears in the derivative. Along a singular arc, we obtain for $q=2$:
$$
(-1)^q\frac{\partial}{\partial\alpha}\frac{d^{2q}}{d\hat{t}^{2q}}\phi(\hat{t})<0,
$$
satisfying the Kelley condition necessary for optimality. Given that the singular arc is of second order, an optimal trajectory can enter into the singular arc only by a \textit{chattering arc} (also called the Fuller's phenomenon, \textit{i.e.}, an arc with an infinite number of switches~\cite{borisov_fullers_2000,marchal_chattering_2013}). \\

@@

The recently published model of Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015} provides a synthesis of the currently available knowledge of the ppGpp regulatory system.
Through the mechanisms of ppGpp production and degradation, it describes regulation of the synthesis of ribosomal RNA.
We explain below how we use the model to compare the action of the ppGpp system with the on-off control strategy.
The denomination of variables and parameters follows the Supporting Information of~\cite{bosdriesz_how_2015} , and is reproduced in Table~\ref{S4.tab:notations} in order to make the text self-contained.

The evolution of the cellular concentration of ppGpp is described in~\cite{bosdriesz_how_2015} by
\begin{equation}
\label{eq:ppGpp}
\frac{d\texttt{ppGpp}}{dt} = v_{RelA}(r_{t,tot}) + v_{spoT} - k_{spoT} \cdot \texttt{ppGpp},
\end{equation}
where $v_{spoT}$ and $k_{spoT}$ are constants (see Table~\ref{S4.tab:notations}), and $v_{RelA}$ is a function of $r_{t,tot}$, the total concentration of "stalled" ribosomes:
\begin{equation}
\label{eq:vrelA}
v_{RelA}(r_{t,tot}) = k_{RelA} \cdot RelA_{tot} \cdot \frac{r_{t,tot}}{K_{D,RelA} + r_{t,tot}}.
\end{equation}
The amount of stalled ribosomes is determined by the equilibrium between charged and uncharged tRNA, $t_{ai}$ and $t_{i}$, in the cell:
\begin{equation}
\label{eq:rttot}
r_{t,tot} = \sum_i r_{ti} = \sum_i r_i \frac{t_i/\kappa_{t}}{1+t_{ai}/\kappa_{ta}+t_i/\kappa_{t}},
\end{equation}
which can be rewritten as
\begin{equation}
\label{eq:rttot-f}
r_{t,tot} = \sum_i r_i \frac{t_i/\kappa_{t}}{1+(0.5 r-t_i)/\kappa_{ta}+t_i/\kappa_{t}},
\end{equation}
using the assumption that $t_{tot,i}  = t_{ai} + t_i = 0.5 \cdot r$.
$r_i$ denotes the concentration of ribosomes recognizing amino acid  $i$.
Finally, with $r = \sum_i r_i$ the total ribosome concentration and $a_i$ the concentration of amino acid $i$, the dynamics of the charged tRNA concentration is described by
\begin{equation}
\label{eq:ti_dynamic}
\frac{dt_{ai}}{dt} = v_{tai}(a_i, t_{i}) - f_i \cdot v_{ribosome}(t_i,r),
\end{equation}
with $v_{tai}(a_i, t_{i})$ the synthesis rate of charged tRNA, and $f_i \cdot v_{ribosome}(t_i,r)$ their consumption via protein synthesis.
In particular,
\begin{eqnarray}
v_{tai}(a_i, t_{i}) &=& k_{Si} \cdot S_{tot,i} \cdot \frac{t_i\, a_i}{t_i\,  K_{Mai} + a_i\,  K_{Mti} + t_i\,  a_i}, \label{eq:vtai}\\
v_{ribosome}(t_i, r) &=& k_{rib} \cdot r \cdot \left(1+ \sum_i \left[ f_i \cdot \left( 1 + \frac{t_i}{\kappa_t} \right) \frac{\kappa_{ta}}{0.5\cdot r - t_i} \right]\right)^{-1}. \label{eq:vrib}
\end{eqnarray}

@@

\begin{equation}
\begin{alignedat}{2}
&A {t_i}^2 + B t_i + C = 0, \;\;\; \text{with}\\
&A = \frac{k_{Si} \,S_{tot,i} \,a_i}{f_i\, k_{rib} \,r} \left( \frac{\kappa_{ta}}{\kappa_t} - 1\right) + K_{Mai} + a_i,\\
&B = \frac{k_{Si} \,S_{tot,i} \,a_i}{f_i \,k_{rib} \,r} (0.5\,r+\kappa_{ta}) + a_i \,K_{Mti} - 0.5\,r\,(K_{Mai} + a_i),\\
&C = - 0.5 \,r \,a_i \,K_{Mti},
\end{alignedat}
\end{equation}
and therefore
\[
t_i (a_i, r) = \frac{-B \pm \sqrt{B^2 - 4AC}}{2A}.
\]
It is not difficult to show that the only solution on $[0, 0.5\,r]$ is
\begin{equation}
\label{eq:ti_final}
t_i (a_i, r) = \frac{-B + \sqrt{B^2 - 4AC}}{2A}.
\end{equation}
From this result, we obtain $r_{t,tot}$ as a function of $a_i$ and $r$, by applying (A1) to Eq.~\ref{S4.eq:rttot-f}:
\begin{equation}
\label{eq:rttot-f_without_sum}
r_{t,tot}(t_i, r) = r \cdot \frac{t_i/\kappa_{t}}{1+(0.5 r-t_i)/\kappa_{ta}+t_i/\kappa_{t}},
\end{equation}
and substituting $t_i$ by the expression of Eq.~\ref{eq:ti_final}.

Finally, we apply (A2) to Eq.~\ref{eq:ppGpp} and obtain the final expression giving the concentration of ppGpp as a function of the total amino acid and ribosome concentrations:
\begin{equation}
\footnotesize
\texttt{ppGpp}(a_i, r) = \frac{1}{k_{spoT}} \left( k_{RelA} \cdot RelA_{tot} \cdot \frac{r_{t,tot}(a_i, r)}{K_{D,RelA} + r_{t,tot}(a_i,r)} + v_{spoT} \right).
\end{equation}
This function is represented in Fig.~\ref{S4.fig:ppGpp} with parameters taken from Table~\ref{tab:notations}.

>>>>

\begin{figure}[tb]
\centering
\includegraphics[clip,trim={0.4cm 0.5cm 0.54cm 0.45cm}]{./Fig/Fig4}
\caption{
\textbf{Optimal control of the self-replicator during a nutrient upshift.}
\textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines. The optimal trajectory is shown as a solid, red curve.
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the switching curve $\varphi(\hat{p})$.
The optimal solution was obtained by numerical optimization using \texttt{bocop}~\cite{bonnans_bocop_2012} (see \textit{Methods} for details), using the parameter values $E_M=1$ and $K=0.003$, and starting from the initial state $(0.024,0.18)$ at $t=0$ (optimal steady state for $E_M=0.2$).
\textit{B:}~Time evolution of the control variable $\alpha_{opt}(\cdot)$ (thick, red line) and the environment $E_M$ (dashed, black line).
}
\label{fig:optimalcontrol}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/Fig6}
\caption{\small
\textbf{Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}
\textit{A:}~Trajectory in the phase plane for the nutrient-only strategy (green curve).
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the $\hat{r}$-nullcline.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with $\alpha = f(E_M)$ as specified by Eq.~\ref{eq:meth_f} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The initial state corresponds to the steady state attained for an environment given by $0.2 \, E_M$.
While converging to the new steady state after the upshift, the precursor concentration makes a large overshoot.
\textit{B:}~As above, but for the precursor-only strategy.
The feedback control strategy is now defined by $\alpha = g(\hat{p})$ as specified by Eq.~\ref{eq:meth_g} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The solution trajectory (blue curve) exhibits a lower overshoot.
\textit{C:}~Evolution of the control variable $\alpha(\cdot)$ as a function of time, for each of the above two strategies.
Notice that in the nutrient-only strategy $\alpha(\cdot)$ immediately jumps to the optimal value for the post-upshift steady state (green curve), whereas in the precursor-only strategy it depends on the (time-varying) precursor concentration (blue curve).
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$  as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:comparison_nutrients_precursor}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/Fig7}
\caption{\small
\textbf{Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}
\textit{A:}~Trajectory in the phase plane for the on-off strategy (yellow curve).
The solid, black curve represents the $\hat{p}$-nullcline and the dashed, black curve the function $g$.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with the equation $\alpha = h(\hat{p},\hat{r})$ defined in Eq.~\ref{eq:stratswitch} and plotted in Fig.~\ref{fig:ppGppsurface}\textit{A}.
The initial state corresponds to the optimal steady state attained for an environment given by $0.2 \, E_M$.
\textit{B:}~Trajectory in the phase plane for the precursor-only strategy (same as in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{B}, added for comparison).
\textit{C:}~Evolution of the control variable $\alpha$ for each strategy as a function of time.
Both strategies stabilize the system at the optimal steady state, but only the on-off strategy (yellow curve) exhibits bang-bang behavior.
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$ for the on-off and precursor-only strategies as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
The final values of $\texttt{Vol} / \texttt{Vol}_{opt}$ attained by the two strategies are 0.9831 and 0.9413, respectively.
The on-off strategy is thus hardly distinguishable from the optimal control strategy in the plot.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:onoffresults}
\end{figure}

@@

As stated in the section \textit{Biomass maximization as an optimal control problem} in the main text, the objective of this study is to maximize the growth rate on an interval $[0,\tau]$ after a nutrient upshift. With Eq.~\ref{eq:supp_growthrate_adim} in \ref{S1_Text}, we have

@@

An extremal trajectory is a quadruplet $(\hat{x}(\cdot),\lambda(\cdot),\lambda_0,\alpha(\cdot))$ satisfying Eqs~\ref{eq:model}-\ref{eq:PMP}. The extremal is said to be normal (resp. abnormal) if $\lambda_0<0$ (resp. $\lambda_0=0$). In the normal case, we normalize the adjoint vector so that $\lambda_0=-1$.

@@

A necessary condition of optimality for a singular arc is given by the Kelley condition~\cite{borisov_fullers_2000}.
We must differentiate $\phi$ with respect to $\hat{t}$ until $\alpha$ appears in the derivative. Along a singular arc, we obtain for $q=2$:
\[
(-1)^q \; \frac{\partial}{\partial\alpha} \; \frac{d^{2q}}{d\hat{t}^{2q}} \; \phi(\hat{t})<0,
\]
satisfying the Kelley condition necessary for optimality. Given that the singular arc is of second order, an optimal trajectory can enter into the singular arc only by a \textit{chattering arc} (also called the Fuller's phenomenon, \textit{i.e.}, an arc with an infinite number of switches~\cite{borisov_fullers_2000,marchal_chattering_2013}).

@@

The recently published model of Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015} provides a synthesis of the currently available knowledge of the ppGpp regulatory system.
Through the mechanisms of ppGpp production and degradation, it describes regulation of the synthesis of ribosomal RNA.
We explain below how we use the model to compare the action of the ppGpp system with the on-off control strategy.
The denomination of variables and parameters follows the Supporting Information of~\cite{bosdriesz_how_2015} , and is reproduced in Table~\ref{tab:notations} in order to make the text self-contained.

The evolution of the cellular concentration of ppGpp is described in~\cite{bosdriesz_how_2015} by
\begin{equation}
\label{eq:ppGpp}
\frac{d\texttt{ppGpp}}{dt} = v_{RelA}(r_{t,tot}) + v_{spoT} - k_{spoT} \cdot \texttt{ppGpp},
\end{equation}
where $v_{spoT}$ and $k_{spoT}$ are constants (see Table~\ref{tab:notations}), and $v_{RelA}$ is a function of $r_{t,tot}$, the total concentration of "stalled" ribosomes:
\begin{equation}
\label{eq:vrelA}
v_{RelA}(r_{t,tot}) = k_{RelA} \cdot RelA_{tot} \cdot \frac{r_{t,tot}}{K_{D,RelA} + r_{t,tot}}.
\end{equation}
The amount of stalled ribosomes is determined by the equilibrium between charged and uncharged tRNA, $t_{ai}$ and $t_{i}$, in the cell:
\begin{equation}
\label{eq:rttot}
r_{t,tot} = \sum_i r_{ti} = \sum_i r_i \frac{t_i/\kappa_{t}}{1+t_{ai}/\kappa_{ta}+t_i/\kappa_{t}},
\end{equation}
which can be rewritten as
\begin{equation}
\label{eq:rttot-f}
r_{t,tot} = \sum_i r_i \frac{t_i/\kappa_{t}}{1+(0.5 r-t_i)/\kappa_{ta}+t_i/\kappa_{t}},
\end{equation}
using the assumption that $t_{tot,i}  = t_{ai} + t_i = 0.5 \cdot r$.
$r_i$ denotes the concentration of ribosomes recognizing amino acid  $i$.
Finally, with $r = \sum_i r_i$ the total ribosome concentration and $a_i$ the concentration of amino acid $i$, the dynamics of the charged tRNA concentration is described by
\begin{equation}
\label{eq:ti_dynamic}
\frac{dt_{ai}}{dt} = v_{tai}(a_i, t_{i}) - f_i \cdot v_{ribosome}(t_i,r),
\end{equation}
with $v_{tai}(a_i, t_{i})$ the synthesis rate of charged tRNA, and $f_i \cdot v_{ribosome}(t_i,r)$ their consumption via protein synthesis.
In particular,
\begin{eqnarray}
v_{tai}(a_i, t_{i}) &=& k_{Si} \cdot S_{tot,i} \cdot \frac{t_i\, a_i}{t_i\,  K_{Mai} + a_i\,  K_{Mti} + t_i\,  a_i}, \label{eq:vtai}\\
v_{ribosome}(t_i, r) &=& k_{rib} \cdot r \cdot \left(1+ \sum_i \left[ f_i \cdot \left( 1 + \frac{t_i}{\kappa_t} \right) \frac{\kappa_{ta}}{0.5\cdot r - t_i} \right]\right)^{-1}. \label{eq:vrib}
\end{eqnarray}

@@

\begin{equation}
\begin{alignedat}{2}
&A {t_i}^2 + B t_i + C = 0, \;\;\; \text{with}\\
&A = \frac{k_{Si} \,S_{tot,i} \,a_i}{f_i\, k_{rib} \,r} \left( \frac{\kappa_{ta}}{\kappa_t} - 1\right) + K_{Mai} + a_i,\\
&B = \frac{k_{Si} \,S_{tot,i} \,a_i}{f_i \,k_{rib} \,r} (0.5\,r+\kappa_{ta}) + a_i \,K_{Mti} - 0.5\,r\,(K_{Mai} + a_i),\\
&C = - 0.5 \,r \,a_i \,K_{Mti},
\end{alignedat}
\end{equation}
and therefore
\[
t_i (a_i, r) = \frac{-B \pm \sqrt{B^2 - 4AC}}{2A}.
\]
It is not difficult to show that the only solution on $[0, 0.5\,r]$ is
\begin{equation}
\label{eq:ti_final}
t_i (a_i, r) = \frac{-B + \sqrt{B^2 - 4AC}}{2A}.
\end{equation}
From this result, we obtain $r_{t,tot}$ as a function of $a_i$ and $r$, by applying (A1) to Eq.~\ref{eq:rttot-f}:
\begin{equation}
\label{eq:rttot-f_without_sum}
r_{t,tot}(t_i, r) = r \cdot \frac{t_i/\kappa_{t}}{1+(0.5 r-t_i)/\kappa_{ta}+t_i/\kappa_{t}},
\end{equation}
and substituting $t_i$ by the expression of Eq.~\ref{eq:ti_final}.

Finally, we apply (A2) to Eq.~\ref{eq:ppGpp} and obtain the final expression giving the concentration of ppGpp as a function of the total amino acid and ribosome concentrations:
\begin{equation}
\texttt{ppGpp}(a_i, r) = \frac{1}{k_{spoT}} \left( k_{RelA} \cdot RelA_{tot} \cdot \frac{r_{t,tot}(a_i, r)}{K_{D,RelA} + r_{t,tot}(a_i,r)} + v_{spoT} \right).
\end{equation}
This function is represented in Fig.~\ref{fig:ppGpp} with parameters taken from Table~\ref{tab:notations}.
